---
title: "2023-03-01_UF_Data_Share"
---

## Data from FLHealthCHARTS Substance Use Dashboard Report

### Years: 2020-2022

URL: [Substance Use Dashboard](https://www.flhealthcharts.gov/Charts/OpenPage.aspx?tn=660)

#### Load packages

```{r}
#| message: false
#| code-fold: true

library(tidyverse)
library(janitor)
library(scales)
```

#### Read data

```{r}
#| code-fold: true

duval_2022 <- readxl::read_xlsx("data/SubstanceUseReport2022_Duval.xlsx", skip = 1) %>%
  clean_names() %>%
  filter(year == "2022")
duval_2021 <- readxl::read_xlsx("data/SubstanceUseReport2021_Duval.xlsx", skip = 1) %>%
  clean_names() %>%
  filter(year == "2021")
duval_2020 <- readxl::read_xlsx("data/SubstanceUseReport2020_Duval.xlsx", skip = 1) %>%
  clean_names() %>%
  filter(year == "2020")
fl_2022 <- readxl::read_xlsx("data/SubstanceUseReport2022_FL.xlsx", skip = 1) %>%
  clean_names() %>%
  filter(year == "2022")
fl_2021 <- readxl::read_xlsx("data/SubstanceUseReport2021_FL.xlsx", skip = 1) %>%
  clean_names() %>%
  filter(year == "2021")
fl_2020 <- readxl::read_xlsx("data/SubstanceUseReport2020_FL.xlsx", skip = 1) %>%
  clean_names() %>%
  filter(year == "2020")


```

#### Clean data

```{r}
#| message: false
#| code-fold: true

# create function to clean data (Duval)

clean_1st_step_duval_fun <- function(df) {
  df %>%
    drop_na() %>%
    mutate(location = "Duval County", .before = indicator) %>%
    pivot_longer(
      -c(location, indicator, measure, year, year_to_date_provisional),
      names_to = "months",
      values_to = "counts"
    ) %>% 
    mutate(counts = if_else(counts == "<5", NA, counts),
           counts = as.numeric(counts))
}

# create function to clean data (FL)

clean_1st_step_fl_fun <- function(df) {
  df %>%
    drop_na() %>%
    mutate(location = "Florida", .before = indicator) %>%
    pivot_longer(
      -c(location, indicator, measure, year, year_to_date_provisional),
      names_to = "months",
      values_to = "counts"
    )
}

# use function on each dataframe

duval_2022 <- clean_1st_step_duval_fun(duval_2022)
duval_2021 <- clean_1st_step_duval_fun(duval_2021)
duval_2020 <- clean_1st_step_duval_fun(duval_2020)
fl_2022 <- clean_1st_step_fl_fun(fl_2022)
fl_2021 <- clean_1st_step_fl_fun(fl_2021)
fl_2020 <- clean_1st_step_fl_fun(fl_2020)

# Combine all into a single dataframe

duval_fl_2020_2022 <- bind_rows(duval_2020, duval_2021, duval_2022, fl_2020, fl_2021, fl_2022)

# remove individual dataframes

rm(duval_2021, duval_2022, fl_2021, fl_2022, duval_2020, fl_2020)

```

## Opioid Data (EMS, ED, Hospitalizations and Overdose Deaths)

### Opioid Overdose Deaths by Year

```{r}
#| code-fold: true

# Overdose deaths by year

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Opioid Overdose Deaths" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Opioid Overdose Deaths",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)

```

### Florida Poison Information Network Calls Related to Opioids

```{r}
#| code-fold: true

# Florida Poison Information Network Calls Related to Opioids

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Florida Poison Information Network Calls Related to Opioids" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Florida Poison Control Calls Related to Opioids",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Emergency Medical Service (EMS) Responses to a Suspected Opioid-involved Overdose by Year

```{r}
#| code-fold: true

# Emergency Medical Service Responses to a Suspected Opioid-involved Overdose

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Emergency Medical Service Responses to a Suspected Opioid-involved Overdose" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of EMS Responses to a\nSuspected Opioid-involved Overdose",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)

```

### Opioid-Involved Non-fatal Overdose Emergency Department (ED) Visits by Year

```{r}
#| code-fold: true

# Opioid-Involved Non-fatal Overdose Emergency Department Visits  

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Opioid-Involved Non-fatal Overdose Emergency Department Visits" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Opioid-Involved Non-fatal Overdose ED Visits",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)

```

### Opioid-involved Non-fatal Overdose Hospitalizations by Year

```{r}
#| code-fold: true

# Opioid-involved Non-fatal Overdose Hospitalizations

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Opioid-involved Non-fatal Overdose Hospitalizations" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Opioid-involved Non-fatal Overdose Hospitalizations",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Heroin-Involved Non-fatal Overdose Emergency Department (ED) Visits by Year

```{r}
#| code-fold: true

# Heroin-Involved Non-fatal Overdose Emergency Department Visits

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Heroin-Involved Non-fatal Overdose Emergency Department Visits" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Heroin-Involved Non-fatal Overdose ED Visits",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Heroin-involved Non-fatal Overdose Hospitalizations by Year

```{r}
#| code-fold: true

# Heroin-involved Non-fatal Overdose Hospitalizations

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Heroin-involved Non-fatal Overdose Hospitalizations" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Heroin-involved\nNon-fatal Overdose Hospitalizations",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

## Opioid Prescription Data

### Number of Prescriptions Dispensed by Year

```{r}
#| code-fold: true

# Number of Prescriptions Dispensed

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Number of Prescriptions Dispensed" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Prescriptions Dispensed",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Number of Unique Patients by Year

```{r}
#| code-fold: true

# Number of Unique Patients by Year

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Number of Unique Patients" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Unique Patients",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Number of Unique Prescribers by Year

```{r}
#| code-fold: true

# Number of Unique Prescribers by Year

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Number of Unique Prescribers" & measure == "Count") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location, scales = "free") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Unique Prescribers",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Prescriptions Dispensed per Patient by Year

```{r}
#| code-fold: true

# Prescriptions Dispensed per Patient by year

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Prescriptions Dispensed per Patient") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location) +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Prescriptions Dispensed per Patient (Avg.)",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```

### Prescriptions Dispensed per Prescriber by Year

```{r}
#| code-fold: true

# Prescriptions Dispensed per Prescriber by year

duval_fl_2020_2022 %>%
  mutate(indicator = str_trim(indicator)) %>% 
  filter(indicator == "Prescriptions Dispensed per Prescriber") %>%
  ggplot(aes(x = indicator, y = year_to_date_provisional, group = year, fill = year)) + 
  geom_col(position = "dodge") +
  geom_text(aes(label = comma(year_to_date_provisional)), position = position_dodge(0.9), vjust = 0) +
  facet_wrap(~location) +
  theme_light() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Number of Prescriptions Dispensed per Prescriber (Avg.)",
       fill = "Year") +
  scale_fill_manual(values=c("#F68E1E",
                             "#00A0B0",
                             "#8DB33E")) +
  scale_y_continuous(labels = comma)
```
